---
title: "session11 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, ggdist, bayestestR, modelr)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
#options(brms.backend = "cmdstanr")
#options(mc.cores = parallel::detectCores())

```

# Introduction

We will be analyzing the data from @roettger2014assessing.

```{r}
#| label: preprocessing

# Load our datasets
incNeutr <- read_csv("../../data/E1_production.csv") |> 
  # we will focus only on the relevant columns
  select(subject, voicing, item_pair, usable, vowel_dur) |> 
  # and only on rows the authors deemed usable
  filter(usable == "usable") |> 
  select(!usable)

```

The variables:

`subject`: unique identifier for participants (n = 16)

`voicing`: underlying voicing status of stop (`voiced` vs. `voiceless`)

`item_pair`: unique identifier for pseudo word minimal pairs (n = 24)

`vowel_dur`: vowel duration in ms

In this study, we recorded people producing words such as Rad and Rat in German and were interested to see whether these words phonetically differ. The reason why this is interesting is that, according to German phonology, these words *should* be identical, i.e. he voicing distinction between /d/ and /t/ should have been *neutralized.* However, previous literature indicates that this neutralization is phonetically *incomplete.* For example, the vowel preceding a (de)voiced stop such /d/ is slightly longer preceding a voiceless stop such as /t/.

The researchers hypothesize that the vowel duration is meaningfully longer preceding voiced stops than preceding voiceless stops. They model this relationship using a linear mixed effects model that accounts for both speaker and word variation as well as by-speaker and by-word variation for the critical predictor voicing.

```{r}
#| label: model_formula

# store model formula
mdl_formula = bf(vowel_dur ~ voicing +
                   # random intercepts and slopes for subject and item_pair
                   (1 + voicing | subject) +
                   (1 + voicing | item_pair))

```

```{r}
#| label: spec-priors

# some weakly informative priors 
my_priors <- c(prior(normal(150,50), class = 'Intercept'),
               # weakly informative prior for the voicing difference (very diffuse)
               prior(normal(0, 30), class = 'b', coef = 'voicingvoiceless'),
               # weakly informative prior for all sources of variance (very diffuse)
               prior(cauchy(0,100), class = 'sd'),
               # weakly informative prior for residual variance (very diffuse)
               prior(cauchy(0,100), class = 'sigma'),
               # correlations
               prior(lkj(1), class = 'cor'))

```

```{r}
#| label: run-model

# run model
xmdl <- brm(mdl_formula,
            family = gaussian(),
            prior = my_priors,
            file = "../../models/in_sesoi",
            data = incNeutr,
            seed = 42)

# I also store the prior-only model 
xmdl_prior <- brm(mdl_formula,
            family = gaussian(),
            prior = my_priors,
            sample_prior = "only",
            file = "../../models/in_sesoi_priors",
            data = incNeutr,
            seed = 42)

```

# Exercises

## (a) Plot the posterior distribution

Explore the posterior distribution for the difference between voiced and voiceless stops. Interpret it.

```{r}

xmdl |> 
  spread_draws(b_voicingvoiceless) |> 
  ggplot(aes(x = b_voicingvoiceless)) +
    stat_halfeye() +
    labs(x = "Posterior distribution for the difference between voiced and voiceless stops")

#Interpretation
#Almost all the posterioir distribution is below zero, meaning that vowels before voiceless stops are shorter that vowels before voiced stops.
```

## (b) Define a SESOI

Define a SESOI and justify it. Can you find an applied justification for a SESOI. Or a theoretical justification? Are there justifications grounded in measurement? What about justifications based on previous research? Define it and formalize it as a Region of Practical Equivalence as we have done before. Use the `bayesfactor_rope()` function to calculate the Bayes factor for and against the posteriors being inside the ROPE.

```{r}

#I believe that a reasonable SESOI here will be 5 ms. I don't have a theoretical justification but I think that it can be grounded in measurement.
rope <- c(-5, 5)

#I had an issue with running the code below, so I tried to (re)install cmdstanr. It didn't help

install.packages(
  "cmdstanr",
  repos = c("https://stan-dev.r-universe.dev", getOption("repos"))
)

library(cmdstanr)

cmdstanr::install_cmdstan()
set_cmdstan_path()

library(cmdstanr)
cmdstanr::cmdstan_version()
cmdstanr::cmdstan_path()


bf_rope <- 
  bayesfactor_rope(
    posterior = xmdl,
    null = rope,
    parameter = "b_voicingvoiceless")

#for some reason the code didn't work, I get this error: Error: Error : Error in require_package("cmdstanr") : 
#Please
#install the
#'cmdstanr' package.
#'#I tried to install it but it fails all the time
#I tried to use chatgpt to see other options but nothing worked. I hope I will fix it at some point, I didn't have any such problems before

bf_rope #can't get a result for it, unfortunatelly

```

## (c) Check robustness of your analysis

How robust is your conclusion against different prior choices?

```{r}

priors_narrow <- c(
  prior(normal(150,50), class = "Intercept"),
  prior(normal(0, 15), class = "b", coef = "voicingvoiceless"),
  prior(cauchy(0,100), class = "sd"),
  prior(cauchy(0,100), class = "sigma"),
  prior(lkj(1), class = "cor"))

priors_wide <- c(
  prior(normal(150,50), class = "Intercept"),
  prior(normal(0, 60), class = "b", coef = "voicingvoiceless"),
  prior(cauchy(0,100), class = "sd"),
  prior(cauchy(0,100), class = "sigma"),
  prior(lkj(1), class = "cor"))
#the models below worked with backend = "rstan"
xmdl_narrow <- brm(
  mdl_formula,
  family = gaussian(),
  prior = priors_narrow,
  data = incNeutr,
  seed = 42,
  backend = "rstan")
#Got the following warning: Warning: Bulk Effective Samples Size (ESS) is too low, indicating posterior means and medians may be unreliable.

xmdl_wide <- brm(
  mdl_formula,
  family = gaussian(),
  prior = priors_wide,
  data = incNeutr,
  seed = 42,
  backend = "rstan")
#no warnings, good

bayesfactor_rope(
  posterior = xmdl_narrow,
  null = rope,
  parameter = "b_voicingvoiceless")
#It worked!

bayesfactor_rope(
  posterior = xmdl_wide,
  null = rope,
  parameter = "b_voicingvoiceless")

#I think that the results are not very robust. With the wide prior BF is close to 1, meaning that the result is unclear. For the narrow one it is 4.26, meaning that the result may be outside the rope. So the result depends on a prior and thus not very robust.

```

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
